---
title: "signifinder R package"
output: 
    github_document:
        toc: false
        toc_depth: 2
        fig_width: 5
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

Signifinder is an R package that collects and implements 46 expression-based signatures from cancer literature. Through the analysis of expression data with the collected signatures, signifinder can attribute to each sample a score per signature that summarizes many different tumor aspects, such as predict the response to therapy or the survival association, as well as quantify multiple microenvironmental conditions, such as hypoxia or the activity of the immune response.

## Installation

You can install the current version of signifinder from GitHub with:

```{r, eval = FALSE, message = FALSE}
BiocManager::install("CaluraLab/signifinder", dependencies = TRUE, build_vignettes = TRUE)
```

<img src=./vignettes/figures/signifinder_visualization.jpg />

