---
title: "signifinder R package"
output: 
    github_document:
        toc: false
        toc_depth: 2
        fig_width: 5
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

`signifinder` is an R package that collects and implements 53 expression-based signatures from cancer literature. Through the analysis of expression data with the collected signatures, `signifinder` can attribute to each sample a score per signature that summarizes many different tumor aspects, such as predict the response to therapy or the survival association, as well as quantify multiple microenvironmental conditions, such as hypoxia or the activity of the immune response.

The `signifinder` package is available from [Bioconductor](https://bioconductor.org/packages/release/bioc/html/signifinder.html), where a vignette containing examples and documentation is included.

## Installation

The `signifinder` package can be installed from Bioconductor:

```{r eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("signifinder")
```

<img src=./vignettes/figures/signifinder_main_figure.png />

